\vspace{-13mm}\includegraphics[width=15mm,scale=3]{../images/IST_Logo.png} \\ {\fontsize{24}{16}Detection of Li in Technological Materials through Nuclear Reactions} \vspace{-5mm}

### 2024-01-20 text/notas/notas.tex
Saturday, 20 January 2024 at 16:31:20

>>>>

\section{Back Scattering - Dúvidas}

o backscattering é todo elástico e um caso particular deste é o de rutherford.
O caso particular de rutherford considera os atomos como esferas perfeitas para calcular a secção eficaz deste fenomeno.
Em geral, as secções eficazes do backscattering obtém-se experimentalmente e não usamos as de rutherford.
O que interessa é que o backscattering elastico como fenomeno e não as suas diferentes secções eficazes.
Pileup - o detetor recebe duas partículas de backscattering em simultaneo e assum que é so uma e por isso soma uma contagem com o dobro da energia.
A particula emitida no backscattering pode ter diferentes valores de energia de acordo com a colisao elastica entre os dois atomos (a mecanica diz nos como calular a energia com que cada átomo, o incidente e o da amosta à superfície, fica depois da colisão) e por isso a soma da energia de duas também pode estar num dado intervalo.
O pico inicial é do ruido eletronico e por isso desconsideramos. Usar detector em vez de channel.
A secção eficaz da reação nuclear com o litio varia com a energia do protão incidente.
O protão incidente perde energia à medida que percorre a amostra devido ao stopping power da amostra em si e por isso a secção eficaz de reacções que ocorrem mais no interior da amostra é diferente das que ocorrem à sua superfície.

### 2024-01-20 text/paper/paper.tex
Saturday, 20 January 2024 at 16:31:20

Before the emission of the reaction products, the reaction leads to an intermediate state, a $\ce{^8Be}$ compound nucleus, which then rapidly decays into two alpha particles that we can detect.% An alpha particle is a helium nucleus, composed of two protons and two neutrons. %toda a gente sabe o que é uma particula alpha valhamedeus

@@

Some of the samples we are going to work with also include boron and according to the beam energy we use it is possible that the protons lead to a nuclear reaction with the boron atoms in the samples. Therefore, the $\ce{^{11}B}$ reaction with a proton is as follows:

\begin{equation}
  \ce{ ^{11}B + ^1H -> ^{12}C -> ^8Be + \alpha}
\label{eq:reactionB}
\end{equation}

The reaction first leads to the formation of a $\ce{^{12}C}$ compound nucleus which decays into one alpha particle and a $\ce{^8Be}$ nucleus. The $\alpha$ particle can be detected with more or less energy according to the energy level on which the $\ce{^8Be}$ nucleus is emitted. If the latter is on the fundamental state, the $\alpha$ particle will have more energy and if it is on a higher state, the $\alpha$ particle will have less energy.

Regarding the phenomena expected in the collected energy spectrum, the primary aspect involves the energy values. Both the alpha particles and other byproducts are emitted with distinct energies immediately following the nuclear reaction. These energies can be calculated based on the reaction and the beam's energy. The computation process will be detailed in section \ref{sec:theoreticalvalues}.
\paragraph{}
Firstly, we expect to see intensity peaks on the spectrum on these specific energies. However, we must take into account the fact that before arriving at the detector, the emitted particles have to go through the other atoms on the sample. As an incident proton comes in, it might react with another proton from the material that sits on its surface but it can also react with one inside it. The byproducts produced inside the sample initially have the energy we compute but do lose part of it as they collide with other particles before reaching the detector. Basically, we must consider the sample's thickness. The bigger the depth at which the reaction takes place, the smaller the energy with whom they are detected because more of it gets dissipated.
Therefore, this will result in an energy profile - the energy peak will be gradually preceded by less intense peaks due to particles that lost energy on their way to the detector. It is less likely for the particles that are emitted inside the sample to reach the detector than for those who were emitted on its surface since on each collision there is a chance that it's direction might change. This is why the intensity on the energy profile has a steady increase before hitting a maximum value at the energy with whom the particle comes out of the reaction.
\paragraph{}
We expect this phenomena to be particularly significant in samples where the atom that takes part in the reaction can be found not only at the surface of the sample but also in its depth. Therefore, in implanted samples where the substrate is only a layer, the energy profile will be less evident.

@@

% o backscattering é todo elástico e um caso particular deste é o de rutherford. 
% O caso particular de rutherford considera os atomos como esferas perfeitas para calcular a secção eficaz deste fenomeno. 
% Em geral, as secções eficazes do backscattering obtém-se experimentalmente e não usamos as de rutherford. 
% O que interessa é que o backscattering elastico como fenomeno e não as suas diferentes secções eficazes. 
% Pileup - o detetor recebe duas partículas de backscattering em simultaneo e assum que é so uma e por isso soma uma contagem com o dobro da energia.
% A particula emitida no backscattering pode ter diferentes valores de energia de acordo com a colisao elastica entre os dois atomos (a mecanica diz nos como calular a energia com que cada átomo, o incidente e o da amosta à superfície, fica depois da colisão) e por isso a soma da energia de duas também pode estar num dado intervalo. 
% O pico inicial é do ruido eletronico e por isso desconsideramos. Usar detector em vez de channel. 
% A secção eficaz da reação nuclear com o litio varia com a energia do protão incidente. 
% O protão incidente perde energia à medida que percorre a amostra devido ao stopping power da amostra em si e por isso a secção eficaz de reacções que ocorrem mais no interior da amostra é diferente das que ocorrem à sua superfície.

@@

In order to choose the energy of the beam we need to take into account the cross section of the reaction between the protons and the Li-7. We collected data from the IBANDL database for this reaction in a range of [0.5,7] MeV and plotted it in the following graph:

\begin{center}
    \label{TT_21}
    \centering
    \includegraphics[scale = 0.17]{../images/Li_crosssection_energy.jpeg}
    \captionof{figure}{Cross section of the reaction between protons and Li-7 as a function of the beam energy}
\end{center}

We are only interested in the energy range of [0.5,5] MeV because the energy of the beam is limited by the accelerators range. As we can see in the above figure, the cross section of the reaction is maximum in the [2.5, 3] MeV range. Therefore, since we want to detect the Li-7 in the samples, we should choose the energy of the beam to be in this range. 
\paragraph{}
However, practically, the accelerator cannot reach a beam within the [2.5, 3] MeV energy range. Increasing beam energies take exponentially more time to be set up. Also, we are interested in the possible boron identification in some of the used samples. Boron does not react with such high energies but a lower energy beam is effective in doing that.
Having this into consideration, we choose a beam energy of $1.3$ MeV for our experimental procedure.

@@

\paragraph{Lithium} Firstly, for the reaction \ref{eq:reactionLi}:
\begin{equation}
\begin{split}
  E_{\alpha_{0_1}} &= 7.630 \text{ MeV} \\
  E_{\alpha_{0_2}} &= 11.031 \text{ MeV} \\
  \theta_{\alpha_{0_1}} &= 165 ^{\circ} \\
  \theta_{\alpha_{0_2}} &= 12.4 ^{\circ} \\
  Q (\ce{^7Li}) &= 17.346 \text{ MeV} \\
\end{split}
\label{eq:energiesli}
\end{equation}

@@

\paragraph{Boron} Secondly, for the reaction \ref{eq:reactionB}:
\begin{equation}
\begin{split}
  E_{\alpha_0} &= 3.749 \text{ MeV} \\
  E_{\ce{^8Be}} &= 3.1116 \text{ MeV} \\
  \theta_{\alpha_0} &= 165 ^{\circ} \\
  \theta_{\ce{^8Be}} &= 11.6 ^{\circ}\\
  Q (\ce{^11Br}) &= 8.591 \text{ MeV} \\
\end{split}
\label{eq:energiesBr}
\end{equation}

We need to comment these energies.

@@

We started by displaying the samples in a vertical sample holder with the given order: in the lowest position the implanted sample, followed by lithium aluminate ($\ce{LiAlO2}$) and finally lithium fluoride ($\ce{LiF}$). In order to do a calibration from the number of counts per channel to energy, we added an additional source on the holder just above the lithium fluoride one. Using a pachymeter, we measured the distance between the samples so that afterwards we could control the hight in order to change the sample on the beam's target.

We placed the sample holder on a support and afterwards both of them inside the chamber. A vacuum machine was connected to the accelerator to reduce the pressure inside both the chamber and the beam.

Afterwards, we started the procedure to turn on the TANDEM accelerator, according to the provided instruction manual. Firstly, we turned on the $\ce{H^+}$ duoplasmatron source of the accelerator as well as the magnets and lenses. Then, we opened the valves to propagate the vacuum throughout the entirety of the tubes.

In the control panel, we adjusted the angles of the lenses and the magnets current according to the manual, so that the beam would reach the samples. Finally, we used a collimator with a 2mm radius to collimate the beam.

Finally, we used a computer program to obtain the spectra of the different sources. The multichannel analyzer (MCA) used to acquire data was gradually set to 70 V and the angle between the top two detectors was 165º.

@@

For the analysis, we began by calculating the theoretical energy values of the two alphas.

DE ONDE VEM ISTO CARAGO? TEMOS DE TER REFERÊNCIAS

@@

The energies obtained were 7629.80 kEv and 11031.44 kEv. (FALTA EXPLICAR PERCENTAGENS)
For the identification of lithium in our samples, we need to compare the theoretical energy values with the experimental ones.
Since we know the emited alpha in our samples is more energetic than the theoretical values, it is expected that the experimental peaks are slightly higher than the theoretical ones.
If the samples contains a peak in the proximity of the alpha peaks it shows that the sample contains lithium.

@@

For the LiF sample, we used the calibration [\ref{sec:calibration}] to convert the channel values into energy values.
we overlapped the expected $\alpha$ energy with the spectrum, and obtained the following:

@@

The experimental $\alpha_0$ is extremely close to the theoretical value, therefore we can confirm the presence of lithium in this sample. 
Moreover, it is according to the expected hypothesis that the observed lithium energy is slightly higher than the theoretical value.
Analysing this spectrum further:

@@

The experimental $\alpha_0$ is very similar to the theoretical value, hence we can confirm the presence of lithium in this sample. 
If we analyze this spectrum further, we can see the following:

@@

There is a peak in the proximity of the $\alpha_0$ peak, which is the expected lithium peak.
One more time we can confirm the presence of lithium in this sample.

If we analyze the spectrum further:

@@

To conclude, we ended up with a successful experiment. We were able to detect the presence of lithium in all the 
analysed samples trough nuclear reactions. Furthermore, we managed to go beyond our proposal and estimate the amount of 
lithium in each sample.
The results obtained were in accordance with the values obtained from the Danish Institute that provided the samples. MEIO INVENTADO

>>>>

Before the emission of the reaction products, the reaction leads to an intermediate state, a $\ce{^8Be}$ compound nucleus, which then rapidly decays into two alpha particles that we can detect.
An alpha particle is a helium nucleus, composed of two protons and two neutrons. % toda a gente sabe o que é uma particula alpha valhamedeus

% \paragraph{}

% Some of the samples we are going to work with also include boron and according to the beam energy we use it is possible that the protons lead to a nuclear reaction with the boron atoms in the samples. Therefore, the $\ce{^{11}B}$ reaction with a proton is as follows:

% \begin{equation}
%   \ce{ ^{11}B + ^1H -> ^{12}C -> ^8Be + \alpha}
% \label{eq:reactionB}
% \end{equation}

% The reaction first leads to the formation of a $\ce{^{12}C}$ compound nucleus which decays into one alpha particle and a $\ce{^8Be}$ nucleus. The $\alpha$ particle can be detected with more or less energy according to the energy level on which the $\ce{^8Be}$ nucleus is emitted. If the latter is on the fundamental state, the $\alpha$ particle will have more energy and if it is on a higher state, the $\alpha$ particle will have less energy.

Regarding the phenomena expected in the collected energy spectrum, the primary aspect involves the energy values.
Both the alpha particles and other byproducts are emitted with distinct energies immediately following the nuclear reaction.
These energies can be calculated based on the reaction and the beam's energy.
The computation process will be detailed in section \ref{sec:theoreticalvalues}.

@@

Firstly, we expect to see intensity peaks on the spectrum on these specific energies.
However, we must take into account the fact that before arriving at the detector, the emitted particles have to go through the other atoms on the sample.
As an incident proton comes in, it might react with another proton from the material that sits on its surface but it can also react with one inside it.
The byproducts produced inside the sample initially have the energy we compute but do lose part of it as they collide with other particles before reaching the detector.
Basically, we must consider the sample's thickness. The bigger the depth at which the reaction takes place, the smaller the energy with whom they are detected because more of it gets dissipated.
Therefore, this will result in an energy profile - the energy peak will be gradually preceded by less intense peaks due to particles that lost energy on their way to the detector.
It is less likely for the particles that are emitted inside the sample to reach the detector than for those who were emitted on its surface since on each collision there is a chance that it's direction might change.
This is why the intensity on the energy profile has a steady increase before hitting a maximum value at the energy with whom the particle comes out of the reaction.

We expect this phenomena to be particularly significant in samples where the atom that takes part in the reaction can be found not only at the surface of the sample but also in its depth.
Therefore, in implanted samples where the substrate is only a layer, the energy profile will be less evident.

@@

@@

In order to choose the energy of the beam we need to take into account the cross-section of the reaction between the protons and the Li-7. We collected data from the IBANDL database for this reaction in a range of [0.5,7] MeV and plotted it in the following graph:

\begin{center}
    \label{TT_21}
    \centering
    \includegraphics[scale = 0.17]{../images/Li_crosssection_energy.jpeg}
    \captionof{figure}{Cross-section of the reaction between protons and Li-7 as a function of the beam energy}
\end{center}

We are only interested in the energy range of [0.5,5] MeV because the energy of the beam is limited by the accelerators range. As we can see in the above figure, the cross-section of the reaction is maximum in the [2.5, 3] MeV range. Therefore, since we want to detect the Li-7 in the samples, we should choose the energy of the beam to be in this range.
\paragraph{}
However, practically, the accelerator cannot reach a beam within the [2.5, 3] MeV energy range. Increasing beam energies take exponentially more time to be set up. Also, we are interested in the possible boron identification in some samples. Boron does not react with such high energies but a lower energy beam is effective in doing that.
Having this into consideration, we choose a beam energy of $1.3$ MeV for our experimental procedure.

@@

\paragraph{Lithium} For the reaction \ref{eq:reactionLi}:
\begin{equation}
\begin{split}
  E_{\alpha_{0_1}} &= 7.62980 \text{ MeV} \\
  E_{\alpha_{0_2}} &= 11.03144 \text{ MeV} \\
  \theta_{\alpha_{0_1}} &= 165 ^{\circ} \\
  \theta_{\alpha_{0_2}} &= 12.4 ^{\circ} \\
  Q (\ce{^7Li}) &= 17.346 \text{ MeV} \\
\end{split}
\label{eq:energiesli}
\end{equation}

@@

% \paragraph{Boron} Secondly, for the reaction \ref{eq:reactionB}:
% \begin{equation}
% \begin{split}
%   E_{\alpha_0} &= 3.749 \text{ MeV} \\
%   E_{\ce{^8Be}} &= 3.1116 \text{ MeV} \\
%   \theta_{\alpha_0} &= 165 ^{\circ} \\
%   \theta_{\ce{^8Be}} &= 11.6 ^{\circ}\\
%   Q (\ce{^11Br}) &= 8.591 \text{ MeV} \\
% \end{split}
% \label{eq:energiesBr}
% \end{equation}

% We need to comment these energies.

@@

We started by displaying the samples in a vertical sample holder with the given order: in the lowest position the implanted sample, followed by lithium aluminate ($\ce{LiAlO2}$) and finally lithium fluoride ($\ce{LiF}$). 
In order to do a calibration we added a triple $\alpha$ source on the holder above the lithium fluoride. 
Using a pachymeter, we measured the distance between the samples so that afterwards we could control the height of the holder in the chamber and change the beam's target.
We placed the sample holder on a support and inserted both of them inside the chamber. 
A vacuum machine was connected to the accelerator to reduce the pressure inside both the chamber and the beam.
Afterwards, we started the procedure to turn on the TANDEM accelerator, according to the provided instruction manual. 
Firstly, we turned on the $\ce{H^+}$ duoplasmatron source of the accelerator as well as the magnets and lenses. 
Then, we opened the valves to propagate the vacuum throughout the entirety of the tubes.
In the control panel, we adjusted the angles of the lenses and the magnets current according to the manual, so that the beam would reach the samples. 
Finally, we used a collimator with a 2mm radius to collimate the beam.
In order to collect the data, we used a computer program to obtain the spectra of the different sources. 
The multichannel analyzer (MCA) used to acquire data was gradually set to 70 V and the angle between the top two detectors was $165 ^{\circ}$.

@@

For the analysis, we began by plotting the theoretical energy values of the two alphas.

@@

The energies obtained were 7629.80 kEv and 11031.44 kEv from section [\ref{sec:theoreticalvalues}].
For the identification of lithium in our samples, we need to compare the theoretical energy values with the experimental ones.
Since we know the emitted alpha particles from our samples are more energetic than their theoretical values, it is expected that the experimental peaks are slightly ahead of the theoretical ones in the spectrum.
If there is a peak in the spectrum that is close to the theoretical value, we can confirm the presence of lithium in the sample.
For the reasons given in section [\ref{sec:theoreticalvalues}], due to the detector's geometry, we expect to only detect $\alpha_0$.

@@

For the LiF sample, we used the calibration [\ref{sec:calibration}] to convert the channel values into energy values.
We overlapped the expected $\alpha$ energy with the spectrum, and obtained the following:

@@

The experimental $\alpha_0$ is extremely close to the theoretical value, therefore we can confirm the presence of lithium in this sample.
Moreover, it is according to the expected hypothesis that the observed lithium energy is slightly higher than the theoretical value.
% Analysing this spectrum further:

@@

The experimental $\alpha_0$ is very similar to the theoretical value, hence we can confirm the presence of lithium in this sample.
% If we analyze this spectrum further, we can see the following:

@@

There is a peak in the proximity of the $\alpha_0$ peak, which is the expected lithium peak.
One more time we can confirm the presence of lithium in this sample.
% If we analyze the spectrum further:

@@

To sum up, we were able to successfully conduct the proposed research project.
Nuclear reactions is a very effective technique to detect the presence of elements in samples, although its cost is high.
We were able to detect the presence of lithium in all the samples. 
Furthermore, we managed to go beyond our proposal and estimate the amount of lithium in each sample.
% The results obtained were in accordance with the values obtained from the Danish Institute that provided the samples. MEIO INVENTADO
